---
title: "Bike Data Analysis - Week 3 |Group By and Probabilities"
author: "Oluwatosin Agbaakin"
date: "2024-01-23"
output: html_document
theme: cerulean
---
<style>
p {
  text-align: justify;
}
</style>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Introduction:

This analysis is on the bike dataset, which focuses on exploring relationships between different variables through grouped data frames, visualization, and hypothesis testing.

I dived deep into individual rows of data, and investigated each row and group, looking into their probability, which can help in anomaly detection. I began by loading the necessary libraries and the dataset. Then, I inspected the dataset to fully understand its structure, variables, and initial statistics.

## Data Loading:

```{r}
library(tidyverse)
bike_data <- read.csv("bike_data.csv")
head(bike_data)
str(bike_data)
summary(bike_data)
```

## Grouping and Summary:

Here, I grouped the data by Marital Status and summarized the Age variable. I aim to understand the age distribution within each marital status category.

```{r}
group_by_marital <- bike_data %>%
  group_by(Marital.Status) %>%
  summarise(Average_Age = mean(Age, na.rm = TRUE))

ggplot(group_by_marital, aes(x = `Marital.Status`, y = Average_Age)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Average Age by Marital Status")
```

From the visualization above, married individuals constitute the larger group (53.8%), while single individuals form the smaller group (46.2%). The average age is slightly higher in the married group compared to the single group.

## **Grouping by Occupation and the Number of children:**

In this section, I analyzed the number of children based on different occupations of the household.

-   The largest group by occupation is the Professional with 27.6%, while the smallest is Manual with 11.9%.

-   The Management group has the highest average number of children, while Manual and Clerical occupations have lower averages.

I visualized the above figures and summary below:

```{r}
group_by_occupation <- bike_data %>%
  group_by(Occupation) %>%
  summarise(Average_Children = mean(Children, na.rm = TRUE))

ggplot(group_by_occupation, aes(x = Occupation, y = Average_Children)) +
  geom_bar(stat = "identity", fill = "tomato") +
  labs(title = "Average Number of Children by Occupation") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## **Grouping by Regions and summarizing Cars:**

This analysis focuses on the number of cars owned in different regions:

-   North America is the largest region with 50.8%, while 'Pacific' has the smallest with 19.2%.

-   On average, people from the Pacific region own more cars compared to the other regions.

```{r}
group_by_region <- bike_data %>%
  group_by(Region) %>%
  summarise(Average_Cars = mean(Cars, na.rm = TRUE))

ggplot(group_by_region, aes(x = Region, y = Average_Cars)) +
  geom_bar(stat = "identity", fill = "brown") +
  labs(title = "Average Number of Cars by Region")
```

## Cross-Tabulation of Categorical Variables:

I further explored the relationship between two categorical variables, 'Marital Status' and 'Occupation', to understand their interaction.

```{r}
marital_occupation_tab <- table(bike_data$`Marital.Status`, bike_data$Occupation)

ggplot(as.data.frame(marital_occupation_tab), aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  labs(title = "Cross Tabulation of Marital Status and Occupation", x = "Marital Status", y = "Occupation")
```

## **Further Analysis:**

-   The smallest probability groups identified (Single for Marital Status, Manual for Occupation, and Pacific for Region) could be subject to further investigation to understand why they are less represented.

-   For instance, the higher number of cars in the Pacific region could be linked to regional characteristics like urbanization, public transport availability, or income levels.

-   Additional analysis can be performed to see how these smallest groups relate to other variables in the dataset, like income or bike purchase decisions.

# Conclusion:

The above heatmap visualization shows the distribution of different occupation categories within marital status groups.

-   It can be clearly observed from the heatmap that the most common occupation for both married and single individuals appears to be Professional as indicated by the darker shade.

-   Clerical occupations are more common among single individuals compared to married individuals.

-   Manual and Management occupations have a moderate representation across both marital statuses.

-   Skilled Manual seems to have a slightly higher frequency among married individuals than single ones, as suggested by the color shading. In relation to the conclusion from the analyzed dataset above:

In relation to the conclusion from the analyzed dataset above:

-   The heatmap visually reinforces the quantitative findings from the data analysis. It shows which occupations are more or less common among different marital statuses.

-   It helps in identifying patterns or trends in occupation with respect to marital status, such as which occupations are predominantly chosen by married or single individuals.

-   The heatmap can support hypotheses regarding socio-economic behavior. For example, if Professional is the most common occupation among both married and single individuals, one might hypothesize that this occupation is a popular choice due to factors like job availability, job desirability, or income levels associated with Professional roles.
